# Polynomial and Splines

##Polynomaials
level 1: 
$$ {Y}_{ij} = \beta_{0j}  + \beta_{1j}(Time_{ij} - \bar{X)} + \beta_{2j}(Time_{ij} - \bar{X)}^2 + \varepsilon_{ij} $$


Level 2: 
$$ {\beta}_{0j} = \gamma_{00} +   U_{0j}$$  
$$ {\beta}_{1j} = \gamma_{10} +  U_{1j} $$ 
$$ {\beta}_{2j} = \gamma_{20} +  U_{2j} $$ 



## polynomial example
```{r}

rm(list = ls())

library(readr)
cdrs <- read_csv("~/Box Sync/5165 Applied Longitudinal Data Analysis/Longitudinal/cdrs.csv")

personality <- read_csv("~/Box Sync/5165 Applied Longitudinal Data Analysis/Longitudinal/Subject_personality.csv")

library(ggplot2)
gg1 <- ggplot(personality,
   aes(x = neodate, y = neuroticism, group = mapid)) + geom_line()  
gg1
```

```{r}
library(tidyverse) 
personality<- personality %>% 
  group_by(mapid) %>%
  arrange(neodate) %>% 
  mutate(wave = seq_len(n())) 
```

```{r}
gg2 <- ggplot(personality,
   aes(x = wave, y = neuroticism, group = mapid)) + geom_line()  
gg2
```


```{r}
personality$neodate <- as.Date(personality$neodate, origin = "1900-01-01")

gg3 <- ggplot(personality,
   aes(x = neodate, y = neuroticism, group = mapid)) + geom_line()  
gg3

```


```{r}
## convert to days from first assessment

personality.wide <- personality %>% 
  dplyr::select(mapid, wave, neodate) %>% 
  spread(wave, neodate) 

personality.wide$wave_1 <- personality.wide$'1'
personality.wide$wave_2 <- personality.wide$'2'
personality.wide$wave_3 <- personality.wide$'3'
personality.wide$wave_4 <- personality.wide$'4'
personality.wide$wave_5 <- personality.wide$'5'

personality.wide <- personality.wide %>% 
mutate (w_1 = (wave_1 - wave_1)/365,
          w_2 = (wave_2 - wave_1)/365,
          w_3 = (wave_3 - wave_1)/365,
          w_4 = (wave_4 - wave_1)/365,
        w_5 = (wave_5 - wave_1)/365)

personality.long <- personality.wide %>% 
  dplyr::select(mapid, w_1:w_5) %>% 
  gather(wave, year, -mapid) %>% 
  separate(wave, c('weeks', 'wave' ), sep="_") %>% 
 dplyr::select(-weeks) 

personality.long$wave <-  as.numeric(personality.long$wave)


personality <- personality %>% 
   left_join(personality.long, by = c('mapid', 'wave' )) 

```


```{r}
gg4 <- ggplot(personality,
   aes(x = year, y = neuroticism, group = mapid)) + geom_line()  
gg4
```



```{r}
library(lme4)

p1 <- lmer(neuroticism ~ year + (1 | mapid), data=personality)
summary(p1)

```




```{r}
library(lme4)
personality.s <- personality %>% 
  group_by(mapid) %>% 
  tally() %>% 
   filter(n >=2) 

 personality <- personality %>% 
   filter(mapid %in% personality.s$mapid)

p2 <- lmer(neuroticism ~ year + (1 | mapid), data=personality)
summary(p2)

```


```{r}
p3 <- lmer(neuroticism ~ year + (year | mapid), data=personality)
summary(p3)
```
### importance of centering
```{r}

personality$year <- as.numeric(personality$year)
  
p4 <- lmer(neuroticism ~ year + I(year^2) + (year | mapid), data=personality)
summary(p4)
# woah, how do I interpret this? WHy all of a sudden non-sig? 
# what would happen if I changed my time metric? 
```

```{r}
library(psych)
describe(personality$year)

personality$year.c <- personality$year - 3.1

p5 <- lmer(neuroticism ~ year.c + I(year.c^2) + (year.c | mapid), data=personality)
summary(p5)
```

### random terms
fitting a random slope plus a random quadratic leads to difficulties ie non-congergence. What does this model say? 
```{r}
p6 <- lmer(neuroticism ~ year + I(year^2) + ( I(year^2) | mapid), data=personality)
summary(p6)
```

## Splines aka piecewise
Fit more than 1 trajectory. Best to use when we have a reason for a qualitative difference at some identified time point. For example, before your health event you may have a different trajectory than after it and thus you would want to model two seperate trajectories. Splines allow you to do this in a single model. You can do this in simple regression and the logic follows for growth models. 

We simply replace time with dummy variables that represent different segments we wish to model. The point of separation is called a knot. You can have as many as you want and these can be pre-specified (usually for our case) or in more advanced treatments have the data specify it for you.   

### seperate curves
The most common is to create different trajectories that change across knots. The easiest example is to take your time variable and transform it into a Time1 and time2, that represent the different time periods. This is easiest to see if we choose our wave variable as our time metric, though you do not have to necessarily do it this way. 


```{r}
t1 <- tribble(
  ~time, ~t0, ~t1,~t2,~t3,~t4,~t5,
  "time 1", 0, 1,2,2,2,2,
  "time 2", 0, 0,0,1,2,3
)
t1
```


The idea is that once you hit the knot your value stays the same. Same logic for the second knot, until you get to that knot you dont have a trajectory. 


This can be contrasted with a different type of coding, called incremental. Here the first trajectory keeps going, whereas the second trajectory starts at the position of the knot. 

```{r}
t2 <- tribble(
  ~time, ~t0, ~t1,~t2,~t3,~t4,~t5,
  "time 1", 0, 1,2,3,4,5,
  "time 2", 0, 0,0,1,2,3
)
t2
```

The two coding schemes propose the same type of trajectoy, the only thing that differes is the interpretation of the coefficients. 

In the first, the two slope coefficients represent the actual slope in the respective time period. 

In the second, the coefficient for time 2 represents the deviation from the slope in period 1. The positive of this second method is you can easily test whether these two slopes are different from one another. 

```{r}

personality$time1 <- recode(personality$wave, '1' = 0 , '2' = 1,  '3' = 1, '4' = 1,'5' = 1)      
personality$time2 <- recode(personality$wave, '1' = 0 , '2' = 0,  '3' = 1, '4' = 2,'5' = 3) 


```


level 1: 

$$ {Y}_{ij} = \beta_{0j}  + \beta_{1j}Time1_{ij}} + \beta_{2j}Time2_{ij} + \varepsilon_{ij} $$




Level 2: 
$$ {\beta}_{0j} = \gamma_{00} +  U_{0j} $$  

$$ {\beta}_{1j} = \gamma_{10} +  U_{1j} $$ 
$$ {\beta}_{2j} = \gamma_{20} +  U_{2j} $$ 



```{r}
p7 <- lmer(conscientiousness ~ time1 + time2 + (time1   | mapid) , data=personality)
summary(p7)
```

```{r}

gg5 <- ggplot(personality, aes(x = wave, y = conscientiousness, group = mapid)) +  stat_smooth(method = 'lm', formula = y ~ poly(x,2, raw = TRUE),data = personality, aes(x = wave, y = conscientiousness, group=1)) + scale_y_continuous(limits = c(30, 40))
gg5


```


### splines + polynomail = polynomial piecewise


$$ {Y}_{ij} = \beta_{0j}  + \beta_{1j}Time1_{ij} +  \beta_{2j}Time1_{ij}^2 + \beta_{3j}Time2_{ij} + \varepsilon_{ij} $$

Level 2: 
$$ {\beta}_{0j} = \gamma_{00} +  U_{0j} $$  

$$ {\beta}_{1j} = \gamma_{10} +  U_{1j} $$ 
$$ {\beta}_{2j} = \gamma_{20} +  U_{2j} $$
$$ {\beta}_{3j} = \gamma_{30} +  U_{3j}$$ 


