# Advanced Multivariate Longitudinal SEM designs 

Thus far our longitudinal SEM designs have mostly dealt with single growth curves. These models assume and underlying trajectory that is unitary across time. In other words, the trajectory may not be linear but it is thought to be resulting from a singular process. What is somewhat unsatisfactory about this is that change in constructs may happen at a different timescale. With growth curves the time scale of change is the entire period. 

An alternative approach is to look at repeated processes of change, thereby getting at more complex theories that can look at a) what construct changes prior to another b) whether changes in a construct occur for the same reasons at different times and c) whether or not the dynamics of change among multiple constructs are similar across time. These more advanced questions necessarily involve breaking the change process down into smaller change quantities. Instead of one growth curve, there may be multiple change (difference) scores. Or there may be multiple regressions associating one construct to another. 

How are these different than the cross-lagged panel model that Josh said was mostly worthless prior in the semester? Well they still aren't great, and that is because it is difficult to address chicken and egg questions that the cross-lag panel model gets used for. However, they do attempt to overcome some of the limitations of those models. Moreover, some of the questions that they attempt to answer side step the difficulties of cross-lagged panel models. 

### Cross-lagged panel model
As an aside, what are the difficulties of cross-lagged panel models? To list a few: 

1. Arbitrary starting point can change association. 
2. Time between lags can influence results because changes may not be aligned with assessment. Think of a wheel rolling and measuring the location of a point at at every quarter rotation compared to 2 rotations. 
3. Different constructs may change at different rates, influencing results. 
4. Theoretically, the model suggest that one point in time influences change in some other construct. Why would Tuesday at 2 be so important, why not Thursday at 4?
5. typically need to constrain cross loadings to be the same to ensure quality estimates. But doing so makes it impossible to see if the association differ across time. 

## Difference scores

The stepping stone to these more advanced models that move beyond cross lagged panel models are built on the humble difference score. 

A long time ago I said that these are not great. Indeed, there is a great history about how difference scores are problematic. This changed the way that experimental psychology analyzed and conducted their experiments in the  60s and early 70s. Cronbach of the alpha fame had a wonderful missive about the lack of utility of difference scores. 

He is right, every difference scores contain measurement error that cannot be separated from true change. In place, people have regressed t2 on t1 and used the residual for a residual change score. However, there are problems such as a loss of information where 2 variables become 1. Low reliability. And there might  be a different factor structure at time 2 from that at time 1.

What to do? Don't use difference scores, use something else. Maybe latently? 

## Analyzing change with two time points

This occurs a lot. Pre-post design. Only have enough money to follow up participants once. Getting an itchy publishing-finger while waiting for your longitudinal study to mature. 

There are a few options (See McArdle Annual Review of Psych for more details). 

The most basic (and most extendable) is really similar to a growth model. To understand where it is coming from: 

(delta T) = T2 - T1
delta T is going to be a latent variable, created by factor loading of 1 from T1 -> T2, meaning that delta T is thought of as an error variance term; what is left over after saying T1 perfectly causes T2. Can now relate that to T1, something the difference score and residual change score models could not do simultaneously. 




```{r, echo=FALSE}
#####  Simulate data for a multiple indicator Univariate Latent Change Score model ####
#Fix sample size
samplesize<-500

#Simulate data for a Univariate Latent Change Score model. 
MILCS_simulate<-'

####    The following two lines specify the measurement model for multiple indicators (X1-X3) 
####    measured on two occasions (T1-T2)
COG_T1=~.8*T1X1+.9*T1X2+.7*T1X3   # This specifies the measurement model for COG_T1 
COG_T2=~.8*T2X1+.9*T2X2+.7*T2X3   # This specifies the measurement model for COG_T2 


#####     The following lines specify the core assumptions of the LCS 
#####     and should not generally be modified

COG_T2 ~ 1*COG_T1           # Fixed regression of COG_T2 on COG_T1
dCOG1 =~ 1*COG_T2           # Fixed regression of dCOG1 on COG_T2
COG_T2 ~ 0*1                # This line constrains the intercept of COG_T2 to 0
COG_T2 ~~ 0*COG_T2          # This fixes the variance of the COG_T2 to 0  


T1X1~0*1                  # This fixes the intercept of X1 to 0
T1X2~0*1                  # This fixes the intercept of X2 to 1
T1X3~.0*1                 # This fixes the intercept of X3 to 0.5 
T2X1~0*1                  # This fixes the intercept of X1 to 0
T2X2~0*1                  # This fixes the intercept of X2 to 1
T2X3~0*1                 # This fixes the intercept of X3 to 0.5


###### The following five parameters will be estimated in the model. 
###### Values can be modified manually to examine the effect on the model

dCOG1 ~ 1.5*1            # This fixes the intercept of the change score to 3 
COG_T1 ~ 6*1           # This fixes the intercept of COG_T1 to 50. 
dCOG1 ~~ .75*dCOG1        # This fixes the variance of the change scores to .75. 
COG_T1 ~~ 1.3*COG_T1      # This fixes the variance of the COG_T1 to 8. 
dCOG1~-0.1*COG_T1       # This fixes the self-feedback parameter to -0.1. 
'

#Simulate data
set.seed(1234)
simdat<-simulateData(MILCS_simulate,sample.nobs = samplesize,meanstructure = T) #Simulate data
colMeans(simdat) #sanity check the means
write.csv(simdat,'2_simdat.csv')
library(tidyverse)
simdat <- as.tibble(simdat)
```





```{r}
change <-

'C_T1=~1*T1X1 + L2*T1X2  + L3*T1X3    
C_T2=~1*T2X1 + L2*T2X2 + L3*T2X3   

C_T2 ~ 1*C_T1     # Fixed regression of C_T2 on C_T1
dC =~ 1*C_T2     # Fixed regression of dC on C_T2
C_T2 ~ 0*1        # This line constrains the intercept of C_T2 to 0
C_T2 ~~ 0*C_T2    # This fixes the variance of the C_T2 to 0 

dC ~ 1             # This estimates the intercept of the change score 
C_T1 ~  1           # This estimates the intercept of C_T1 
dC ~~  dC       # This estimates the variance of the change scores 
C_T1 ~~ C_T1    # This estimates the variance of the COG_T1 
dC ~ C_T1          # This estimates the self-feedback parameter, which for now is just the covaration with change and T1


T1X1~~T2X1   # This allows residual covariance on indicator X1 across T1 and T2
T1X2~~T2X2   # This allows residual covariance on indicator X2 across T1 and T2
T1X3~~T2X3   # This allows residual covariance on indicator X3 across T1 and T2

T1X1~~T1X1   # This allows residual variance on indicator X1 
T1X2~~T1X2   # This allows residual variance on indicator X2
T1X3~~T1X3   # This allows residual variance on indicator X3

T2X1~~T2X1  # This allows residual variance on indicator X1 at T2 
T2X2~~T2X2  # This allows residual variance on indicator X2 at T2 
T2X3~~T2X3  # This allows residual variance on indicator X3 at T2

T1X1~0*1                 # This constrains the intercept of X1 to 0 at T1
T1X2~M2*1                   # This estimates the intercept of X2 at T1
T1X3~M3*1                   # This estimates the intercept of X3 at T1
T2X1~0*1                 # This constrains the intercept of X1 to 0 at T2
T2X2~M2*1   # This estimates the intercept of X2 at T2
T2X3~M3*1   # This estimates the intercept of X3 at T2


'

change.fit <- sem(change, data=simdat, missing="ML")
summary(change.fit, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)
```


This gives you everything you need for two time points. How would you introduce a predictor variable? Multiple group model? Bi-variate latent change model? 


## multi wave latent change model

The real benefit of this type of framework is building on this type of change model to incorporate multiple change models simultanously. This allows one to look at multiple change processes in a single model, opening the door to analyzing more dynamic processes. One can fit two latent change scores (LCSs) with 3 repeated assessments, 3LCSs with 4 repeated measures, etcetera. 

This is similar to the Steyer Model's that Leah will present. 


One top of this type of model you can fit a growth curve. Why?  Can capture both general trends and more fine-grained temporal dynamics. This can be especially useful when trying to separate a known, more stable change occurring during development (e.g.natural agining process) from more specific fluctuations due to external factors (e.g., life events).

```{r, echo = FALSE}
BD_simulate<-'

COGlv_T1=~1*COG_T1        # Defining the COG latent variables
COGlv_T2=~1*COG_T2        # Defining the COG latent variables
COGlv_T3=~1*COG_T3        # Defining the COG latent variables
COGlv_T4=~1*COG_T4        # Defining the COG latent variables

NEUlv_T1=~1*NEU_T1        # Defining the NEU latent variables
NEUlv_T2=~1*NEU_T2        # Defining the NEU latent variables
NEUlv_T3=~1*NEU_T3        # Defining the NEU latent variables
NEUlv_T4=~1*NEU_T4        # Defining the NEU latent variables

##### The following parameters capture the core assumptions of the LCS and should not generally be modified

COGlv_T2 ~ 1*COGlv_T1     # This parameter regresses COG_T2 perfectly on COG_T1
COGlv_T3 ~ 1*COGlv_T2     # This parameter regresses COG_T3 perfectly on COG_T2
COGlv_T4 ~ 1*COGlv_T3     # This parameter regresses COG_T4 perfectly on COG_T3

NEUlv_T2 ~ 1*NEUlv_T1     # This parameter regresses NEU_T2 perfectly on NEU_T1
NEUlv_T3 ~ 1*NEUlv_T2     # This parameter regresses NEU_T3 perfectly on NEU_T2
NEUlv_T4 ~ 1*NEUlv_T3     # This parameter regresses NEU_T4 perfectly on NEU_T3

dCOG1 =~ 1*COGlv_T2       # This defines the change score as measured perfectly by scores on COG_T2
dCOG2 =~ 1*COGlv_T3       # This defines the change score as measured perfectly by scores on COG_T3
dCOG3 =~ 1*COGlv_T4       # This defines the change score as measured perfectly by scores on COG_T4

dNEU1 =~ 1*NEUlv_T2       # This defines the change score as measured perfectly by scores on NEU_T2
dNEU2 =~ 1*NEUlv_T3       # This defines the change score as measured perfectly by scores on NEU_T3
dNEU3 =~ 1*NEUlv_T4       # This defines the change score as measured perfectly by scores on NEU_T4

COG_T1~~5*COG_T1          # This specifies the COG residual variances 
COG_T2~~5*COG_T2          # This specifies the COG residual variances 
COG_T3~~5*COG_T3          # This specifies the COG residual variances 
COG_T4~~5*COG_T4          # This specifies the COG residual variances 

NEU_T1~~5*NEU_T1          # This specifies the NEU residual variances 
NEU_T2~~5*NEU_T2          # This specifies the NEU residual variances 
NEU_T3~~5*NEU_T3          # This specifies the NEU residual variances 
NEU_T4~~5*NEU_T4          # This specifies the NEU residual variances 

#Dynamics

dNEU1~0.05*NEUlv_T1       # This specifies the NEU self-feedback parameter (equality constrained across timepoints)
dNEU2~0.05*NEUlv_T2       # This specifies the NEU self-feedback parameter (equality constrained across timepoints) 
dNEU3~0.05*NEUlv_T3       # This specifies the NEU self-feedback parameter (equality constrained across timepoints)

dCOG1~0.05*COGlv_T1       # This specifies the COG self-feedback parameter (equality constrained across timepoints)
dCOG2~0.05*COGlv_T2       # This specifies the COG self-feedback parameter (equality constrained across timepoints) 
dCOG3~0.05*COGlv_T3       # This specifies the COG self-feedback parameter (equality constrained across timepoints)


dNEU1~.3*COGlv_T1         # This specifies the COG to NEU coupling parameter 
dNEU2~.3*COGlv_T2         # This specifies the COG to NEU coupling parameter 
dNEU3~.3*COGlv_T3         # This specifies the COG to NEU coupling parameter 

dCOG1~0.4*NEUlv_T1        # This specifies the NEU to COG coupling parameter 
dCOG2~0.4*NEUlv_T2        # This specifies the NEU to COG coupling parameter 
dCOG3~0.4*NEUlv_T3        # This specifies the NEU to COG coupling parameter 


iCOG=~1*COGlv_T1                   # This defines the COG intercept measurement model
sCOG=~1*dCOG1+1*dCOG2+1*dCOG3      # This defines the COG slope measurement model
iCOG~2*1                           # This specifies the COG intercept intercept (mean)
iCOG~~2*iCOG                       # This specifies the COG intercept variance
sCOG~2*1                           # This specifies the COG slope intercept
sCOG~~3*sCOG                       # This specifies the COG slope variance

iNEU=~1*NEUlv_T1                   # This defines the NEU slope measurement model
sNEU=~1*dNEU1+1*dNEU2+1*dNEU3      # This defines the NEU slope measurement model
iNEU~2*1                           # This specifies the NEU intercept intercept (mean)
iNEU~~2*iNEU                       # This specifies the NEU intercept variance
sNEU~2*1                           # This specifies the NEU slope intercept
sNEU~~3*sNEU                       # This specifies the NEU slope variance

iNEU~~.8*sNEU                      # This specifies the iNEU sNEU covariance
iNEU~~.8*sCOG                      # This specifies the iNEU sCOG covariance
iNEU~~.8*iCOG                      # This specifies the iNEU iCOG covariance
iCOG~~.8*sCOG                      # This specifies the iCOG sCOG covariance        
iCOG~~.8*sNEU                      # This specifies the iCOG sNEU covariance
sCOG~~.8*sNEU                      # This specifies the sCOG sNEU covariance  

'

set.seed(1234)
simdatBD<-simulateData(BD_simulate,sample.nobs = samplesize,meanstructure = T) #Simulate data
colMeans(simdatBD) 

write.csv(simdatBD,'simdatBD.csv')


```


```{r}

BDCS<-'

COGlv_T1=~1*COG_T1        # Defining the COG latent variables
COGlv_T2=~1*COG_T2        # Defining the COG latent variables
COGlv_T3=~1*COG_T3        # Defining the COG latent variables
COGlv_T4=~1*COG_T4        # Defining the COG latent variables

NEUlv_T1=~1*NEU_T1        # Defining the NEU latent variables
NEUlv_T2=~1*NEU_T2        # Defining the NEU latent variables
NEUlv_T3=~1*NEU_T3        # Defining the NEU latent variables
NEUlv_T4=~1*NEU_T4        # Defining the NEU latent variables


COGlv_T2 ~ 1*COGlv_T1     # This parameter regresses COG_T2 perfectly on COG_T1
COGlv_T3 ~ 1*COGlv_T2     # This parameter regresses COG_T3 perfectly on COG_T2
COGlv_T4 ~ 1*COGlv_T3     # This parameter regresses COG_T4 perfectly on COG_T3

NEUlv_T2 ~ 1*NEUlv_T1     # This parameter regresses NEU_T2 perfectly on NEU_T1
NEUlv_T3 ~ 1*NEUlv_T2     # This parameter regresses NEU_T3 perfectly on NEU_T2
NEUlv_T4 ~ 1*NEUlv_T3     # This parameter regresses NEU_T4 perfectly on NEU_T3

dCOG1 =~ 1*COGlv_T2       # This defines the change score as measured perfectly by scores on COG_T2
dCOG2 =~ 1*COGlv_T3       # This defines the change score as measured perfectly by scores on COG_T3
dCOG3 =~ 1*COGlv_T4       # This defines the change score as measured perfectly by scores on COG_T4

dNEU1 =~ 1*NEUlv_T2       # This defines the change score as measured perfectly by scores on NEU_T2
dNEU2 =~ 1*NEUlv_T3       # This defines the change score as measured perfectly by scores on NEU_T3
dNEU3 =~ 1*NEUlv_T4       # This defines the change score as measured perfectly by scores on NEU_T4

COG_T1~~COG_T1          # This estimates the COG residual variances 
COG_T2~~COG_T2          # This estimates the COG residual variances 
COG_T3~~COG_T3          # This estimates the COG residual variances 
COG_T4~~COG_T4          # This estimates the COG residual variances 

NEU_T1~~NEU_T1          # This estimates the NEU residual variances 
NEU_T2~~NEU_T2          # This estimates the NEU residual variances 
NEU_T3~~NEU_T3          # This estimates the NEU residual variances 
NEU_T4~~NEU_T4          # This estimates the NEU residual variances 

#Dynamics

dNEU1~B1*NEUlv_T1       # This estimates the NEU self-feedback parameter 
dNEU2~B1*NEUlv_T2       # This estimates the NEU self-feedback parameter 
dNEU3~B1*NEUlv_T3       # This estimates the NEU self-feedback parameter 

dCOG1~B2*COGlv_T1       # This estimates the COG self-feedback parameter 
dCOG2~B2*COGlv_T2       # This estimates the COG self-feedback parameter 
dCOG3~B2*COGlv_T3       # This estimates the COG self-feedback parameter 

dNEU1~C1*COGlv_T1         # This estimates the COG to NEU coupling parameter 
dNEU2~C1*COGlv_T2         # This estimates the COG to NEU coupling parameter 
dNEU3~C1*COGlv_T3         # This estimates the COG to NEU coupling parameter 

dCOG1~C2*NEUlv_T1        # This estimates the NEU to COG coupling parameter 
dCOG2~C2*NEUlv_T2        # This estimates the NEU to COG coupling parameter 
dCOG3~C2*NEUlv_T3        # This estimates the NEU to COG coupling parameter 


iCOG=~1*COGlv_T1                   # This defines the COG intercept measurement model
sCOG=~1*dCOG1+1*dCOG2+1*dCOG3      # This defines the COG slope measurement model
iCOG~1                             # This estimates the COG intercept intercept (mean)
iCOG~~iCOG                         # This estimates the COG intercept variance
sCOG~1                             # This estimates the COG slope intercept
sCOG~~sCOG                         # This estimates the COG slope variance

iNEU=~1*NEUlv_T1                   # This defines the NEU slope measurement model
sNEU=~1*dNEU1+1*dNEU2+1*dNEU3      # This defines the NEU slope measurement model
iNEU~1                             # This estimates the NEU intercept intercept (mean)
iNEU~~iNEU                         # This estimates the NEU intercept variance
sNEU~1                             # This estimates the NEU slope intercept
sNEU~~sNEU                         # This estimates the NEU slope variance

iNEU~~sNEU                      # This estimates the iNEU sNEU covariance
iNEU~~sCOG                      # This estimates the iNEU sCOG covariance
iNEU~~iCOG                      # This estimates the iNEU iCOG covariance
iCOG~~sCOG                      # This estimates the iCOG sCOG covariance        
iCOG~~sNEU                      # This estimates the iCOG sNEU covariance
sCOG~~sNEU                      # This estimates the sCOG sNEU covariance  

'

fitBDCS <- lavaan(BDCS, data=simdatBD, missing="ML")
summary(fitBDCS, fit.measures=TRUE, standardized=TRUE) 
```



## ALT and ALT-SR

ANother way to build both a growth model and look at more specific time associations. 



	
